Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-08-28 TACO-paper-v2/coffee.tex
Rewrite full application study

The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve. Once computed, the local element matrices and vectors are respectively inserted in a ``global'' sparse matrix and a global vector; these form a linear system, which is usually solved with an iterative method (the ``solve'' phase). The amount of time spent in assembly and solve varies from problem to problem. As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g. because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what instead impact the solve are, for instance, polynomial order and mesh discretization). The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e. additional known functions that characterize the equation). We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$). More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already in case 1).

The application runs on a tetrahedral mesh composed of 196608 elements. In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$. To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.

@@

In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, split over two figures (one for each polynomial order), without and with optimizations enabled. A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), in solving the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions). The \emph{non-opt} and \emph{opt} bars correspond, respectively, to the cases in which no optimizations and a combination of optimizations have been applied. The optimizations applied are generalized loop-invariant code motion, alignment and padding, and expression splitting. The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.

We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, there is only a minimal performance gain. This changes instantly moving to $p=2$ and $f=2$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}. Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).

In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$. However, it is notably important noting that, as we anticipated, full-application speed ups rise proportionally with the time spent in assembly and, therefore, with the complexity of the equation. Further, by increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant.

@@

In Firedrake, COFFEE has the fundamental role of decoupling the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from platform-specific code optimization. Although at the COFFEE layer there is no notion of finite element method, it is clear that the construction of the compiler and the code transformations were dictated and inspired by our domain of interest. The starting point of our work was a mathematical formulation of a local assembly operation, which is expressible as follows

\begin{equation*}
\scriptsize
\forall_{i, j} (A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) 
\end{equation*}

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$, computing the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ depend on the specific problem being solved and can be intricately complex, involving for example the evaluation of derivatives. Translating into code such a formulation of local assembly is a challenging process for complicated equations; for this reasons, research in automated code generation has been performed over the last few years. Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
\begin{enumerate}
\item All quantities are known values (apart obviously from the unknown $A$). This means that ``good'' code can be generated, for example with explicit loop bounds (i.e. no need to leave them as parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE. 
\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure lies at the core of generalized loop-invariant code motion. Whenever sub-expressions invariant with respect to different sets of loops can be found, the question of whether, where and how to hoist them arises. Any codes presenting these characteristics is amenable to this kind of transformation. 
\end{enumerate}

@@

>>>>

The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve. The evaluation of all local element matrices and vectors, and their insertion, respectively, in a ``global'' sparse matrix and a global vector, compose the ``assembly'' phase. The global matrix and vector form a linear system, usually solved by an iterative method; this is the ``solve'' phase. The percentage of time spent in assembly and solve varies from problem to problem. As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the \textit{complexity} of the partial differential equation (e.g. because of larger loops and more articulated expressions), while the solve time increases with polynomial order and mesh discretization. The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e. additional known functions that characterize the equation).

We study two versions of the static linear elasticity problem: 1) with one coefficient function, $f=1$, and 2) with two coefficient functions, $f=2$. More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already with $f=1$. For each of these two versions, we examine the cases of polynomial order $p=1$ and $p=2$. To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}. The application runs on a tetrahedral mesh composed of 196608 elements.

@@

In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, split over two figures (one for each polynomial order), without and with optimizations enabled. A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), in solving the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions). The \emph{non-opt} and \emph{opt} bars correspond, respectively, to the cases in which no optimizations and a combination of optimizations have been applied. The optimizations applied are generalized loop-invariant code motion, alignment and padding, and expression splitting. We recall that the cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.

We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved. This changes instantly moving to $p=2$ and $f=2$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}. Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).

In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$. However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation. By increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant. The choice of studying the static linear elasticity equation was to show that even relatively simple problems can be characterized by a large proportion of execution time spent in assembly.

@@

In Firedrake, COFFEE has the fundamental role of decoupling the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from platform-specific code optimization. Although at the COFFEE layer there is no notion of finite element method, it is clear that the construction of the compiler and the code transformations were dictated and inspired by our domain of interest. The starting point of our work was the mathematical formulation of a local assembly operation, which is expressible as follows

\begin{equation*}
\scriptsize
\forall_{i, j} ~~ (~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) ~)
\end{equation*}

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$, computing the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ depend on the specific problem being solved and can be intricately complex, involving for example the evaluation of derivatives. Translating into code such a formulation of local assembly is a challenging process for complicated equations; for this reasons, research in automated code generation has been performed over the last few years. Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
\begin{enumerate}
\item All quantities are known values (apart, obviously, from the unknown $A$). This means that ``good'' code can be generated, for example with explicit loop bounds (i.e. no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE. 
\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure lies at the core of generalized loop-invariant code motion. Whenever sub-expressions invariant with respect to different sets of loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises. Any codes presenting these characteristics is amenable to this kind of transformation. Note that, in general, for applying our technique the loop nest could be arbitrarily complex (e.g. more loops), provided that it is still affine.
\item 
\end{enumerate}

@@

%TODO: from the expression tree to a decorated AST.
% Diderot, tensor contraction, image processing?
